Mixture Models for Single Cell Assays

### 2012-03-16 MIMOSA.tex
Updating write-up of multivariate MIMOSA
Signed-off-by: Greg Finak <gfinak@fhcrc.org>

\documentclass[11pt]{article}
\usepackage{geometry,authblk,multicol,amsmath}
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage[textsize=small]{todonotes}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}

@@

\title{Mixture Models for Single Cell Assays}
\author[1]{Greg Finak}
\author[1]{\ldots Others \ldots}
\author[1]{Raphael Gottardo}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
\date{\today}

@@

\subsection{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
Positivity calls for vaccine responders and non--responders depend upon the selection of an appropriate threshold. Therefore, to compare different methods of analysis, comparable thresholds for positivity must be selected for the methods. Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines \todo[inline]{IS THIS CORRECT, OR IS IT WITHIN INDIVIDUALS ACROSS STIMULATIONS?}. In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one--sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.

@@

>>>>

\documentclass[11pt]{article}
\usepackage{geometry,authblk,multicol,amsmath}
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage[textsize=small]{todonotes}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{booktabs}

@@

\title{Mixture Models for Single Cell Assays}
\author[1]{Greg Finak}
\author[2]{SC De Rosa}
\author[1]{Raphael Gottardo}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, WA}
\affil[2]{HIV Vaccine Trials Network, Fred Hutchinson Cancer Research Center (FHCRC), Seattle, Wa}
\date{\today}

@@

\subsection{Statistical Analysis of Responder and Non--responder Calls for Direct Comparison Against the Bayesian Mixture Model Approach}
Positivity calls for vaccine responders and non--responders depend upon the selection of an appropriate threshold. Therefore, to compare different methods of analysis, comparable thresholds for positivity must be selected for the methods. Our mixture modelling approach is fit within each stimulation, and we make positivity calls based on a false discovery rate calculated across individuals, within each stimulation, whereas the originally published analysis makes multiple testing adjustments within individuals, across cytokines. In order to have comparable response rates, we reanalyzed the ICS data using Fisher's one--sided exact test (as described in the original publication) but made positivity calls based on the false discovery rate computed across individuals within each stimulation.

@@

\subsection*{Detecting coexpression via the Multinomial-Dirichlet model}
We applied the Multinomial--Dirichlet extension of MIMOSA to the HVTN054 trial data to identify observations with cells expressing polyfunctional cytokine profiles. As a proof of principle we limited the model to the two--component case (Figure~\ref{tab:nesting}), with one component for the null model with no difference between stimulations, and one component for the alternative model with a difference between stimulations. We fit the model to the IL4 and TNFa cytokines for Env1 stimulated CD4 T--cells from the day 28 time point. The classification of responder and non--responder individuals was compared between the multivariate MIMOSA model, univariate MIMOSA models applied to each cytokine combination, and Fisher's exact test applied to each cytokine combination.

The multivariate MIMOSA model detected two additional responders than Fisher's exact test at the 1\% FDR level (Table~\ref{tab:MDFisher, top left.}. The univariate MIMOSA model applied to each cytokine combination individually also outperformed Fisher's exact test, detecting one more responder for the IL4 and TNFa combination (Table~\ref{tab:MDFisher} bottom right), and two more responders for the TNFa and not IL4 cytokine combination (Table~\ref{tab:MDFisher} bottom left).  In the multivariate case, the two responders detected by MIMOSA that were missed by Fisher's exact test had a  2.4--fold and 2.5--fold increase in the proportion of TNFa+/IL4- cells in the stimulated samples than the unstimulated samples. These are the same two samples detected by the univariate MIMOSA model for TNFa+/IL4- cells, one of which also exhibits a 1.7--fold increase in IL4+/TNFa- expression between stimulated and unstimulated samples that was detected by univariate MIMOSA, but not detected by Fisher's exact test.

\begin{table}[ht]
\begin{center}
\begin{minipage}{0.9\linewidth}
\caption{Confusion matrices comparing responders and non--responders identified by the MIMOSA models and Fisher's exact test. Top left: Fisher's exact test vs the multivariate MIMOSA model for any combination of cytokines.Top right: Fisher's exact test vs MIMOSA for the univariate combination of IL4 and not TNFa. Bottom left: Fisher's exact test vs univariate MIMOSA for TNFa and not IL4. Bottom right: Fisher's exact test vs univariate MIMOSA for IL4 and TNFa. False discovery rate thresholds were set at 1\% for each cytokine combination (except in the multivariate model, where it was 1\% for each observation). }\label{tab:MDFisher}
\end{minipage}
\begin{minipage}{0.49\linewidth}
\center
\begin{tabular}{rrr}
  \hline
  \multicolumn{3}{c}{Any Combination}\\
  \hline
&\multicolumn{2}{c}{MIMOSA}\\
\cmidrule(r){2-3}
 & NR & R \\ 
\cmidrule(r){1-1}
Fisher's Exact Test&&\\
  \hline
NR &  32 &   2 \\ 
  R &   0 &   4 \\ 
   \hline
\end{tabular}
\end{minipage}
\begin{minipage}{0.49\linewidth}
\center
\begin{tabular}{rrr}
\hline
\multicolumn{3}{c}{IL4 and not TNFa}\\
  \hline
&\multicolumn{2}{c}{MIMOSA}\\
\cmidrule(r){2-3}
 & NR & R \\ 
\cmidrule(r){1-1}
Fisher's Exact Test&&\\
  \hline
NR &  38 &   0 \\ 
  R &   0 &   0 \\ 
   \hline
\end{tabular}
\end{minipage}

\begin{minipage}{0.49\linewidth}
\center
\begin{tabular}{rrr}
  \hline
  \multicolumn{3}{c}{TNFa and not IL4}\\
  \hline
&\multicolumn{2}{c}{MIMOSA}\\
\cmidrule(r){2-3}
 & NR & R \\ 
\cmidrule(r){1-1}
Fisher's Exact Test&&\\
  \hline
NR &  36 &   2 \\ 
  R &   0 &   0 \\ 
   \hline
\end{tabular}
\end{minipage}
\begin{minipage}{0.49\linewidth}
\center
\begin{tabular}{rrr}
  \hline
  \multicolumn{3}{c}{IL4 and TNFa}\\
  \hline
&\multicolumn{2}{c}{MIMOSA}\\
\cmidrule(r){2-3}
 & NR & R \\ 
\cmidrule(r){1-1}
Fisher's Exact Test&&\\
  \hline
NR &  33 &   1 \\ 
  R &   0 &   4 \\ 
   \hline
\end{tabular}
\end{minipage}
\end{center}
\end{table}
